Informing policy via dynamic models: Eliminating cholera in Haiti

### 2022-09-02 ms.Rnw
Minor changes
Created new file to fit model 1
refactored replication of lee et al model 1

%%%% parameters %%%%%%%%%%%
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\missing{\hspace{6mm} ---}
\newcommand\fixed{\color{blue}}
\newcommand\demography{D}
\newcommand\code[1]{\texttt{#1}}
\newcommand\paramVec{\theta}

@@

@@

The latent state is described as $\bm{X}^{(3)}(t) = \big(S_{u\vaccCounter}(t),\allowbreak I_{u\vaccCounter}(t),\allowbreak A_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter k}(t),\allowbreak W_u(t),\allowbreak u \in \seq{0}{U},\allowbreak \vaccCounter \in \seq{0}{4},\allowbreak k \in \seq{1}{3}\big)$.
Here, $\vaccCounter=0$ corresponds to unvaccinated, $\vaccCounter=2j-1$ corresponds to a single dose on the $j$th vaccination campaign in unit $u$ and $\vaccCounter=2j$ corresponds to receiving two doses on the $j$th vaccination campaign.
$k\in\seq{1}{3}$ models non-exponential duration in the recovered class before waning of immunity.
The force of infection is
\begin{eqnarray}
\label{eq:model3:foi}
\lambda_u(t) &=& \Wbeta{_u} \frac{W_u(t)}{\Wsat+2W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv 0}(t)+ \asymptomRelativeInfect A_{\varv 0}(t)\big)
\\[-3pt]
\label{eq:model3:SI}
\mu_{S_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \, d\Gamma/dt
\\[-3pt]
\label{eq:model3:SA}
\mu_{S_{u\vaccCounter}A_{u\vaccCounter}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \,  d\Gamma/dt
\\[-3pt]
\label{eq:model3:IR}
\mu_{I_{u\vaccCounter}R_{u\vaccCounter 1}} &=& \mu_{A_{u\vaccCounter}R_{u\vaccCounter 1}} = \muIR
\\[-3pt]
\label{eq:model3:IS}
\mu_{I_{u\vaccCounter}S_{u0}} &=& \muDeath + \choleraDeath
\\[-3pt]
\label{eq:model3:AS}
\mu_{A_{u\vaccCounter}S_{u0}} &=& \muDeath
\\[-3pt]
\label{eq:model3:RRnext}
\mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS
\\[-3pt]
\label{eq:model3:RS}
\mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}}
\\[-3pt]
\label{eq:model3:water}
\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J(t))^r \big] D_i \, \Wshed \big[ I_{u0}(t)+ \asymptomRelativeShed A_{u0}(t) \big]
\\[-3pt]
\label{eq:model3:Decay}
\mu_{W_u\demography} &=& \Wremoval
\end{eqnarray}

@@

The use of likelihood-based inference allows for the statistical testing of potential improvements to the model.
% TODO: We should address somewhere why we included the process noise in Model 1. Anna suggested this below:
% We first demonstrate this capability by proposing the inclusion of additional stochasticity in the latent process through the force of infection, similar to that in Model~3. Eq.~\myeqref{model1:lambda} becomes
% \begin{equation}
% \label{model1:lambda_wn}
% \lambda(t) = \Big(\sum_{k=0}^{K}  I_{k}(t) + \asymptomRelativeInfect \sum_{k=0}^{K} A_{k}(t) \Big)^\nu d\Gamma_u(t)/dt \transmission(t)/N
% \end{equation}
% where $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise.
%
% Part of the appeal of mechanistic models lies in their ability to illustrate a dynamical system.
% Inherent is dynamical systems is randomness, and, in the context of epidemiology, this randomness is a crucial component to the system itself.
% Useful and statistically sound incidence forecasting and prediction of vaccine efficacy are not only dependent upon a reasonable reflection of the natural system in the model; it is necessary for the model to also reflect the quantitative characteristics of the processes in question.
% This need, itself, is dependent upon inclusion of sufficient stochasticity in the model \citep{breto09}.
% For now, we motivate this decision by its presence in Model~3 and leave the more rigorous assessment of this adjustment to the supplement.
% The above can no doubt be improved.
% \arc{I think that evaluating the inclusion of both the noise and the trend parameter in the main text may distract from the main point of this manuscript, which is why I mentioned putting it in the supplement.}
% \eic{The dynamic noise issue is important, but is well described elsewhere so can perhaps be handled mostly by references. Maybe we don't need to do more than mention that Lee et al constrained the noise parameter to be zero, but we found considerable advantage from including the parameter (XXX log units of likelihood) consistent with previous investigations (refs).}
% Transition phrase here?
We demonstrate this capability by proposing a linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
\begin{equation}
\label{model1:betat}
\log\transmission(t) = \sum_{j=1}^6 \transmission_s s_j(t) + \transmissionTrend\bar{t}
\end{equation}
Where $\bar{t}$ is the linear mapping $\bar{t}: [0, N] \rightarrow [-1, 1]$ of the time $t$.
The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
% The inclusion of a possible trend in transmission is due to observing an apparent continual decrease in reported cholera infections from 2012-2019 can be seen in Fig.~\ref{CholeraData}.
While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \citep{rebaudet19CATI}, may have greatly reduced cholera transmission \citep{rebaudet21}.

@@

<<Table 2 Lee Input, echo=FALSE, include=FALSE>>=
load('model1/output/lee_mod1_epidemic_fit.rda')
mod1_lee_epi_ll <- max(prof_if$loglik, na.rm = TRUE)
rm(prof_if)
gc()
load('model1/output/lee_mod1_endemic_fit.rda')
mod1_lee_end_ll <- max(prof_if$loglik, na.rm = TRUE)
rm(prof_if)
gc()

mod1_lee_ll <- mod1_lee_epi_ll + mod1_lee_end_ll
mod1_lee_n_params <- 20
mod1_lee_aic <- 2 * mod1_lee_n_params - 2 * mod1_lee_ll

@@

Immunity (year)

@@

Birth/death (year)

@@

Asympt.~infectivity

@@

Human to human

@@

Water to human

@@

      \hspace{-1mm} {\fixed \Wsat=1}

@@

Human to water

@@

Water survival (wk)

@@

Proc. noise (wk$^{1/2}$)

@@

  & \eqref{model1:meas}

@@

  & \eqref{model2:meas}

@@

  & \eqref{model3:meas}

@@

Obs. overdispersion

@@

  & \eqref{model1:meas}

>>>>

%%%% parameters %%%%%%%%%%%
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\missing{---}
\newcommand\fixed{\color{blue}}
\newcommand\demography{D}
\newcommand\code[1]{\texttt{#1}}
\newcommand\paramVec{\theta}

@@

<<ReplicateModel1, echo=FALSE, message=FALSE, include=FALSE>>=
# In this chunk, we replicate the results in Lee et al for model 1.
# We do this replication here in the ms.Rnw file because we need to get
# estimates for the likelihood of the models used by lee et al. The chunk has
# the following steps:
#   (1) Initialize some starting values close to those used in lee et al
#   (2) do small MIF, similar to what was done in lee et al, to get a fit for
#       epidemic period.
#   (3) Simulate from model
#   (4) Use simulations to initialize endemic phase of the model, and
#       perform MIF, similar to what lee et al did for this phase
#   (5) Simulate from model
#   (6) Calculate the likelihood of each possible set of parameters.

N_PARAMS   <- switch(RUN_LEVEL,  10,  20,  300)
N_LEE_SIMS <- switch(RUN_LEVEL,  10,  20,   20)
NMIF       <- switch(RUN_LEVEL,   5,  20,   50)
NP_MIF     <- switch(RUN_LEVEL,  20,  50,  100)  # Too few, but matches Lee et al
NP_EVAL    <- switch(RUN_LEVEL,  50, 200, 2000)
NREPS_EVAL <- switch(RUN_LEVEL,   3,   5,   36)

#-------------------------- Step 1 ----------------------

set.seed(141837)
sample_beta_nu <- function(mu_beta = 5.8, mu_nu = 0.96, n_params = N_PARAMS) {
  # This function is used to sample starting values for beta and nu.
  #
  # params:
  #   mu_beta: Average of the marginal distribution for beta1. Value was fit visually.
  #   mu_nu: Average of the marginal distribution for nu. Value was fit visually.
  # returns:
  #   data.frame of size 300x2, values of beta1 and nu with truncated bivariate
  #   normal relation.

  # Determine a good value for the marginal standard deviations
  sd_beta <- (12.1-mu_beta) / qnorm(0.995)
  sd_nu <- (0.9-mu_nu) / qnorm(0.005)

  # Create covariance matrix for the bivariate-normal distribution
  cov_mat <- rbind(
    c(sd_beta^2, -0.7 * sd_beta * sd_nu),
    c(-0.7 * sd_beta * sd_nu, sd_nu^2)
  )

  # Sample from the bivariate normal distribution
  X <- MASS::mvrnorm(n = n_params, mu = c(mu_beta, mu_nu), Sigma = cov_mat)
  colnames(X) <- c("beta1", "nu")

  # Identify values that fell outside of range, so they can be resampled
  bad_beta <- which(X[, 'beta1'] > 12.1 | X[, 'beta1'] < 0.04)
  bad_nu <- which(X[, 'nu'] > 1 | X[, 'nu'] < 0.9)
  n_resamp <- length(unique(c(bad_beta, bad_nu)))

  # Keep resampling values until all values fall within the range.
  while (n_resamp > 0) {
    to_resamp <- unique(c(bad_beta, bad_nu))
    X[to_resamp, ] <- MASS::mvrnorm(n = n_resamp, mu = c(mu_beta, mu_nu), Sigma = cov_mat)

    bad_beta <- which(X[, 'beta1'] > 12.1 | X[, 'beta1'] < 0.04)
    bad_nu <- which(X[, 'nu'] > 1 | X[, 'nu'] < 0.9)
    n_resamp <- length(unique(c(bad_beta, bad_nu)))
  }

  # Return results as a data.frame
  as.data.frame(X)
}
# Run the function to get values of beta1 and nu
starting_params <- sample_beta_nu()

sample_rho <- function(beta, n_params = N_PARAMS) {
  # This function is used to get starting values for the parameter rho, based
  # on it's strong correlation with beta1 and nu. Here, we first sample beta1
  # and nu, and then sample rho based on a regression on beta1. Additionally,
  # it is noted that rho has a bi-modal distribution, with lower values of rho
  # corresponding with high probability to higher values of beta1.
  #
  # params:
  #   beta: previously sampled values of beta1 that will be used to get values
  #         of rho.
  # returns:
  #   vector of length 300 that have an appropriate relationship with the parameter
  #   beta1.

  rhos <- rep(0, N_PARAMS)

  # Get slope and intercept for the regression
  slope <- (0.4/(5.1 - 6.75))
  intercpt <- -6.75 * slope

  # Determine whice mode should the value of rho belong to, based on regression
  case_when(
    beta > 6.75 ~ 0,  # Large beta --> group "0"
    # Small beta --> more likely to be group "1", with increasing probability as beta gets smaller
    beta <= 6.75 & beta > (0.95 - intercpt) / slope ~ slope * beta + intercpt,
    TRUE ~ 0.95
  ) -> probs

  # Sample the groups that rho belongs to
  rho_group1 <- rbernoulli(N_PARAMS, probs)
  rho_group2 <- !rho_group1

  # Keep track of how many values are in each group, used later for resampling
  n_group1 <- sum(rho_group1)
  n_group2 <- sum(rho_group2)

  # Resample values of rho that are larger than 1
  group1 <- rnorm(n_group1, mean = 1, sd = 0.05)
  n_bad1 <- sum(group1 > 1)
  while(n_bad1 > 0) {
    group1[group1 > 1] <- rnorm(n_bad1, mean = 1, sd = 0.175)
    n_bad1 <- sum(group1 > 1)
  }

  # get standard deviation for regression
  std_beta1 <- (beta[rho_group2] - mean(beta[rho_group2])) / sd(beta[rho_group2])

  group2 <- 0.25 - 0.025 * std_beta1 - (rlnorm(n_group2, sdlog = 0.2) - 1)

  # Resample values of rho in group2 as needed
  which_bad2 <- group2 < .13
  while(sum(which_bad2) > 0) {
    group2[group2 < 0.13] <- 0.25 - 0.1 * std_beta1[which_bad2] - (rlnorm(sum(which_bad2), sdlog = 0.2) - 1)

    which_bad2 <- group2 < .13
  }

  rhos[rho_group1] <- group1
  rhos[rho_group2] <- group2

  # Return the values of rho
  rhos
}
# Sample rho, using already sampled beta1.
starting_params$rho <- sample_rho(starting_params$beta1)

sample_tau <- function(mu = 4.1, n_params = N_PARAMS) {
  # Function used to sample tau. Tau is approximately independent of the other
  # parameters.
  #
  # params:
  #   mu: mean of the truncated normal distribution
  # returns:
  #   vector of length 300, sampled values of tau from truncated normal distribution

  # Sample from normal distribution
  X <- rnorm(n_params, mean = mu, sd = (1.7-mu) / qnorm(0.001))

  # Get how many resamples are needed
  n_resamp <- sum(X > 5.9 | X < 1.7)

  # Keep resampling until all values fall within desired range
  while (n_resamp > 0) {
    X[X > 5.9 | X < 1.7] <- rnorm(n_resamp, mean = mu, sd = (1.7-mu) / qnorm(0.001))

    n_resamp <- sum(X > 5.9 | X < 1.7)
  }

  # Return values of tau
  X
}
# Use function to sample tau
starting_params$tau <- sample_tau()

# Here we note that there are additional parameters that were fit, but their
# relationship between the remaining parameters is not given in Lee et al. (2020).
# Because of this, we simply sample these parameters using a truncated normal distribution:

POP <- 10911819
E_0_probs <- dnorm(seq(1, 1752, 1), mean = 400, sd = (1752-400) / qnorm(0.99))
I_0_probs <- dnorm(seq(1, 2271, 1), mean = 500, sd = (2271-500) / qnorm(0.99))
starting_params$E_0 <- sample(1:1752, size = N_PARAMS, replace = TRUE, prob = E_0_probs) / POP
starting_params$I_0 <- sample(1:2271, size = N_PARAMS, replace = TRUE, prob = I_0_probs) / POP
starting_params$S_0 <- 1 - starting_params$E_0 - starting_params$I_0

rm(sample_beta_nu, sample_rho, sample_tau, POP, I_0_probs, E_0_probs)
gc()

#-------------------------- Step 2 ----------------------

# Create epi model for haiti1
h1_epi <- haiti1()

bounds <- tribble(
  ~param, ~lower, ~upper,
  "beta2", 1e-08, 10,
  "beta3", 1e-08, 10,
  "beta4", 1e-08, 10,
  "beta5", 1e-08, 10,
  "beta6", 1e-08, 10
)

lower <- bounds$lower
names(lower) <- bounds$param

upper <- bounds$upper
names(upper) <- bounds$param

betas <- sobol_design(
  lower = lower,
  upper = upper,
  nseq  = N_PARAMS
)

epi_params <- cbind(starting_params, betas)

not_in_start_params <- names(coef(h1_epi))[!names(coef(h1_epi)) %in% colnames(epi_params)]

for (i in 1:length(not_in_start_params)) {
  epi_params[, not_in_start_params[i]] <- coef(h1_epi)[not_in_start_params[i]]
}

rm(i)

epi_params$sig_sq <- 0

# simply re-ordering to match model order
epi_params <- epi_params[, names(coef(h1_epi))]

epi_rw <- rw.sd(
  # Parameters close to desired values
  beta1 = 0.001,
  tau   = 0.001,
  nu    = 0.001,
  rho   = 0.001,
  # Parameters not close to desired values
  beta2 = 0.02,
  beta3 = 0.02,
  beta4 = 0.02,
  beta5 = 0.02,
  beta6 = 0.02,
  E_0   = ivp(0.1),
  I_0   = ivp(0.1)
)

registerDoRNG(48963587)

lee1_epi_mif <- bake(
  file = paste0("model1/", rl_dir, "lee1_epi_fit.rds"),
  {
    foreach(
      i = 1:nrow(epi_params),
      .packages = c('pomp'),
      .combine = c
    ) %dopar% {
      r_params <- unlist(epi_params[i, ])
      coef(h1_epi) <- r_params
      mif2(
        h1_epi,
        Np = NP_MIF,
        Nmif = NMIF,
        cooling.fraction.50 = 0.5,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
        rw.sd = epi_rw
      )
    }
  }
)

rm(epi_rw, epi_params, not_in_start_params,
   lower, upper, betas, bounds, starting_params)
gc()

# epi_complete <- as.data.frame(t(coef(lee1_epi_mif)))
# epi_complete$logLik <- logLik(lee1_epi_mif)

# GGally::ggpairs(
#   filter(epi_results, logLik > -2200),
#   columns = c("rho", "tau", "beta1", "nu", "logLik")
# )

# epi_complete <- filter(epi_complete, logLik > -2200)

lee_epi_ll <- bake(
  file = paste0("model1/", rl_dir, "lee1_epi_evals.rds"), {

    # Create data.frame to save results
    lee_epi_ll <- data.frame(
      "pfLL" = rep(0, length(lee1_epi_mif)),
      "pfse" = rep(0, length(lee1_epi_mif))
    )

    for (j in 1:nrow(lee_epi_ll)) {
      # Get parameters of interest
      pf_params <- coef(lee1_epi_mif[[j]])

      # make results reproducible
      registerDoRNG((j * 687383921) %% 7919)

      # Calculate log likelihoods
      ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
        logLik(pfilter(h1_epi, params = pf_params, Np = NP_EVAL))
      }
      lee_epi_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]
      lee_epi_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]
    }

    lee_epi_ll
  }
)

# First, we need to simulate the epidemic model using the fitted parameters.
registerDoRNG(18599687)
foreach(
  i = 1:length(lee1_epi_mif),
  .combine = rbind
) %dopar% {
  guess <- coef(lee1_epi_mif)
  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
                   params = guess)  # 20 sims for each set of parameters
  sims$param_set <- i
  sims
} -> all_sims

#-------------------------- Step 3 ----------------------

# quants <- all_sims %>%
#   select(.id, week, cases) %>%
#   group_by(week) %>%
#   summarize(
#     q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
#     q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
#     q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
#   ) %>%
#   ungroup() %>%
#   mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))

#-------------------------- Step 4 ----------------------

h1_end <- haiti1(period = "endemic")
start_states <- all_sims %>% filter(week == max(week))

end_rw <- rw.sd(
  beta1 = 0.05,
  tau   = 0.005,
  nu    = 0.04,
  rho   = 0.01,
  beta2 = 0.05,
  beta3 = 0.05,
  beta4 = 0.05,
  beta5 = 0.05,
  beta6 = 0.05
)

registerDoParallel(20)
registerDoRNG(21489587)

end_start <- t(coef(lee1_epi_mif))

# end_start <- epi_complete %>%
#   select(-logLik)

lee1_end_mif <- bake(
  file = paste0("model1/", rl_dir, "lee1_end_fit.rds"),
  {
    foreach(
      i = 1:nrow(end_start),
      .packages = c('pomp'),
      .combine = c
    ) %dopar% {

      output <- list()
      r_params <- end_start[i, ]

      start_states %>%
        filter(param_set == i) %>%
        select(-.id, -week, -param_set) %>%
        sapply(., median, na.rm = TRUE) -> rinit_parms

      # Set the inital state
      new_rinit <- Csnippet(
        sprintf(
          "
          S = nearbyint(%f);
          E = nearbyint(%f);
          I = nearbyint(%f);
          A = nearbyint(%f);
          R = nearbyint(%f);
          incid = nearbyint(%f);
          foival = %f;
          Str0 = nearbyint(%f);
          Sout = nearbyint(%f);
          Sin = nearbyint(%f);
          ",
          rinit_parms['S'],
          rinit_parms['E'],
          rinit_parms['I'],
          rinit_parms['A'],
          rinit_parms['R'],
          rinit_parms['incid'],
          rinit_parms['foival'],
          rinit_parms['Str0'],
          rinit_parms['Sout'],
          rinit_parms['Sin']
        )
      )

      h1_end <- h1_end %>% pomp(
        rinit = new_rinit,
        statenames = c(
          "S", "E", "I", "A", "R", "incid",
          "foival", "Str0", "Sout", "Sin"
        )
      )

      coef(h1_end) <- r_params
      mif2(
        h1_end,
        Np = NP_MIF,
        Nmif = NMIF,
        cooling.type = "hyperbolic",
        cooling.fraction.50 = 0.05,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
        rw.sd = end_rw
      )
      # output[['m2']] <- m2
      #
      # coef(h1_end) <- coef(m2)
      # sims <- simulate(  # Step 5
      #   h1_end, nsim = N_LEE_SIMS, format = 'data.frame',
      #   params = coef(m2)
      # )
      # sims$param_set <- i
      #
      # output[['sims']] <- sims
      # output
    }
  }
)

rm(end_rw, start_states)
gc()

lee_end_ll <- bake(
  file = paste0("model1/", rl_dir, "lee1_end_evals.rds"), {

    # Create data.frame to save results
    lee_end_ll <- data.frame(
      "pfLL" = rep(0, length(lee1_end_mif)),
      "pfse" = rep(0, length(lee1_end_mif))
    )

    # Perform particle filter to get likelihood estimates
    for (j in 1:nrow(lee_end_ll)) {
      # Get parameters of interest
      pf_params <- coef(lee1_end_mif[[j]])

      # make results reproducible
      registerDoRNG((j * 687383921) %% 7919)

      # Calculate log likelihoods
      ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
        logLik(pfilter(h1_end, params = pf_params, Np = NP_EVAL))
      }

      lee_end_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]
      lee_end_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]
    }

    lee_end_ll
  }
)

rm(lee1_end_mif, lee1_epi_mif)
gc()

# MIF_params <- matrix(nrow = length(lee1_end_mif$m2) / 2, ncol = length(coef(lee1_end_mif[[1]])))
# all_sims_end <- data.frame()
# liks <- rep(0, nrow(end_start))
# for (i in 1:nrow(end_start)) {
#   MIF_params[i, ] <- coef(lee1_end_mif[[(2 * i) - 1]])
#   all_sims_end <- rbind(all_sims_end, lee1_end_mif[[(2 * i)]])
#   liks[i] <- logLik(lee1_end_mif[[(2 * i) - 1]])
# }
#
# colnames(MIF_params) <- names(coef(lee1_end_mif[[1]]))
# end_results <- as.data.frame(MIF_params)
# end_results$logLik <- liks

# GGally::ggpairs(
#   filter(end_results, logLik > -2200),
#   columns = c("rho", "tau", "beta1", "nu", "logLik")
# )

#-------------------------- Step 6 ----------------------

mod1_lee_ll <- max(lee_end_ll$pfLL, na.rm = TRUE) + max(lee_epi_ll$pfLL, na.rm = TRUE)

rm(
  lee_end_ll, lee_epi_ll, h1_end, h1_epi, all_sims, end_start,
  N_LEE_SIMS, N_PARAMS, NMIF, NP_EVAL, NP_MIF, NREPS_EVAL
)
gc()
@

@@

The latent state is described as $\bm{X}^{(3)}(t) = \big(S_{u\vaccCounter}(t),\allowbreak I_{u\vaccCounter}(t),\allowbreak A_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter k}(t),\allowbreak W_u(t),\allowbreak u \in \seq{0}{U},\allowbreak \vaccCounter \in \seq{0}{4},\allowbreak k \in \seq{1}{3}\big)$.
Here, $\vaccCounter=0$ corresponds to unvaccinated, $\vaccCounter=2j-1$ corresponds to a single dose on the $j$th vaccination campaign in unit $u$ and $\vaccCounter=2j$ corresponds to receiving two doses on the $j$th vaccination campaign.
$k\in\seq{1}{3}$ models non-exponential duration in the recovered class before waning of immunity.
The force of infection is
\begin{eqnarray}
\label{eq:model3:foi}
\lambda_u(t) &=& \Wbeta{_u} \frac{W_u(t)}{1+W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv 0}(t)+ \asymptomRelativeInfect A_{\varv 0}(t)\big)
\\[-3pt]
\label{eq:model3:SI}
\mu_{S_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \, d\Gamma/dt
\\[-3pt]
\label{eq:model3:SA}
\mu_{S_{u\vaccCounter}A_{u\vaccCounter}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \,  d\Gamma/dt
\\[-3pt]
\label{eq:model3:IR}
\mu_{I_{u\vaccCounter}R_{u\vaccCounter 1}} &=& \mu_{A_{u\vaccCounter}R_{u\vaccCounter 1}} = \muIR
\\[-3pt]
\label{eq:model3:IS}
\mu_{I_{u\vaccCounter}S_{u0}} &=& \muDeath + \choleraDeath
\\[-3pt]
\label{eq:model3:AS}
\mu_{A_{u\vaccCounter}S_{u0}} &=& \muDeath
\\[-3pt]
\label{eq:model3:RRnext}
\mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS
\\[-3pt]
\label{eq:model3:RS}
\mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}}
\\[-3pt]
\label{eq:model3:water}
\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J(t))^r \big] D_i \, \Wshed \big[ I_{u0}(t)+ \asymptomRelativeShed A_{u0}(t) \big]
\\[-3pt]
\label{eq:model3:Decay}
\mu_{W_u\demography} &=& \Wremoval
\end{eqnarray}

@@

The ability to maximize the likelihood allows for likelihood-based inference, like performing statistical tests for potential improvements to the model.
% TODO: We should address somewhere why we included the process noise in Model 1. Anna suggested this below:
% We first demonstrate this capability by proposing the inclusion of additional stochasticity in the latent process through the force of infection, similar to that in Model~3. Eq.~\myeqref{model1:lambda} becomes
% \begin{equation}
% \label{model1:lambda_wn}
% \lambda(t) = \Big(\sum_{k=0}^{K}  I_{k}(t) + \asymptomRelativeInfect \sum_{k=0}^{K} A_{k}(t) \Big)^\nu d\Gamma_u(t)/dt \transmission(t)/N
% \end{equation}
% where $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise.
%
% Part of the appeal of mechanistic models lies in their ability to illustrate a dynamical system.
% Inherent is dynamical systems is randomness, and, in the context of epidemiology, this randomness is a crucial component to the system itself.
% Useful and statistically sound incidence forecasting and prediction of vaccine efficacy are not only dependent upon a reasonable reflection of the natural system in the model; it is necessary for the model to also reflect the quantitative characteristics of the processes in question.
% This need, itself, is dependent upon inclusion of sufficient stochasticity in the model \citep{breto09}.
% For now, we motivate this decision by its presence in Model~3 and leave the more rigorous assessment of this adjustment to the supplement.
% The above can no doubt be improved.
% \arc{I think that evaluating the inclusion of both the noise and the trend parameter in the main text may distract from the main point of this manuscript, which is why I mentioned putting it in the supplement.}
% \eic{The dynamic noise issue is important, but is well described elsewhere so can perhaps be handled mostly by references. Maybe we don't need to do more than mention that Lee et al constrained the noise parameter to be zero, but we found considerable advantage from including the parameter (XXX log units of likelihood) consistent with previous investigations (refs).}
% Transition phrase here?
We demonstrate this capability by proposing a linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
\begin{equation}
\label{model1:betat}
\log\transmission(t) = \sum_{j=1}^6 \transmission_s s_j(t) + \transmissionTrend\bar{t}
\end{equation}
Where $\bar{t}$ is the linear mapping $\bar{t}: [0, N] \rightarrow [-1, 1]$ of the time $t$.
The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
% The inclusion of a possible trend in transmission is due to observing an apparent continual decrease in reported cholera infections from 2012-2019 can be seen in Fig.~\ref{CholeraData}.
While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \citep{rebaudet19CATI}, may have greatly reduced cholera transmission \citep{rebaudet21}.

@@

<<Table 2 Lee Input, echo=FALSE, include=FALSE>>=
mod1_lee_n_params <- 20
mod1_lee_aic <- 2 * mod1_lee_n_params - 2 * mod1_lee_ll

@@

\text{Immunity} \\
\text{(year)}

@@

Birth/death (yr)

@@

\text{Asymptomatic} \\
\text{infectivity}

@@

\text{Human to}\\
\text{human}\end{array}$

@@

\text{Water to}\\ \text{human}

@@

@@

\text{Human to}\\
\text{water}\end{array}$

@@

\text{Water} \\
\text{survival (wk)}

@@

\text{Process} \\
\text{noise} (\text{wk}^{1/2})

@@

  & (S16) % TODO: Make sure manual equations match!

@@

  & (S17)

@@

  & (S18)

@@

\text{Observation} \\
\text{overdispersion}

@@

  & (S16)
